---
title: "AI Portfolio Optimization"
author: "Monet Knupp"
date: "2026-04-19"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    toc_collapsed: 
      # collapsed: true
      # smooth_scroll: true
    theme: simplex
  pdf_document:
    toc: true
---
<style>
p {text-align: justify;} h1.title { text-align: font-weight: bold; } 
</style>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1ST AI Prompt
"Can you write R code using quantmod to download adjusted closing prices from Yahoo Finance for the following ETFs over the last 3 years: SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, and GLD. Please make sure to organize the code clearly and include package installation commands."

[ChatGPT](https://chatgpt.com/share/69e53ea2-e08c-83ea-bc5c-443398df8c46)


```{r}
############################################################
# Download Adjusted Closing Prices for ETFs from Yahoo Finance
# ETFs: SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, GLD
# Period: Last 3 Years
############################################################

# -----------------------------
# Setup: Install & Load Packages
# -----------------------------

required_packages <- c("quantmod", "xts", "PerformanceAnalytics", "corrplot")

# Set CRAN mirror (THIS FIXES YOUR ERROR)
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install missing packages only
installed_packages <- rownames(installed.packages())

for (pkg in required_packages) {
  if (!(pkg %in% installed_packages)) {
    install.packages(pkg)
  }
}

# Load libraries
library(quantmod)
library(xts)
library(PerformanceAnalytics)
library(corrplot)

# -----------------------------
# 3. Define ETF tickers
# -----------------------------
etf_symbols <- c("SPY", "QQQ", "IWM", "VEA", "VWO",
                 "AGG", "TLT", "LQD", "VNQ", "GLD")

# -----------------------------
# 4. Define date range
# -----------------------------
end_date <- Sys.Date()
start_date <- end_date - 365 * 3

# -----------------------------
# 5. Download price data
# -----------------------------
getSymbols(Symbols = etf_symbols,
           src = "yahoo",
           from = start_date,
           to = end_date,
           auto.assign = TRUE)

# -----------------------------
# 6. Extract Adjusted Close prices
# -----------------------------
adjusted_prices_list <- lapply(etf_symbols, function(symbol) {
  Ad(get(symbol))
})

# Rename each column to just the ticker symbol
for (i in seq_along(adjusted_prices_list)) {
  colnames(adjusted_prices_list[[i]]) <- etf_symbols[i]
}

# Merge all adjusted close series into one object
adjusted_prices <- do.call(merge, adjusted_prices_list)

# -----------------------------
# 7. View results
# -----------------------------
head(adjusted_prices)
tail(adjusted_prices)

# Optional: check structure
str(adjusted_prices)

# -----------------------------
# 8. Optional - Save to CSV
# -----------------------------
adjusted_prices_df <- data.frame(Date = index(adjusted_prices),
                                 coredata(adjusted_prices))

write.csv(adjusted_prices_df,
          file = "etf_adjusted_prices_last_3_years.csv",
          row.names = FALSE)

# -----------------------------
# 9. Optional - Plot prices
# -----------------------------
plot(adjusted_prices,
     main = "Adjusted Closing Prices of Selected ETFs (Last 3 Years)",
     cex.main = 1.2,     # controls title size (smaller = cleaner)
     adj = 0.5)          # centers the title
```

**A slightly cleaner version for just the main task is this:**

```{r cars}
# Install packages if needed
pkgs <- c("quantmod", "xts")
new_pkgs <- pkgs[!(pkgs %in% rownames(installed.packages()))]
if (length(new_pkgs) > 0) install.packages(new_pkgs)

# Load libraries
library(quantmod)
library(xts)

# ETF symbols
symbols <- c("SPY", "QQQ", "IWM", "VEA", "VWO",
             "AGG", "TLT", "LQD", "VNQ", "GLD")

# Date range: last 3 years
start_date <- Sys.Date() - 365 * 3
end_date <- Sys.Date()

# Download data
getSymbols(symbols, src = "yahoo", from = start_date, to = end_date)

# Extract adjusted close prices
prices <- do.call(merge, lapply(symbols, function(sym) Ad(get(sym))))
colnames(prices) <- symbols

# Preview
head(prices)
tail(prices)
```

### 2ND. AI Prompt

great! using the downloaded ETF price data for SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, and GLD, write R code to calculate daily log returns, annualized returns, annualized volatility, and a correlation matrix. Also include a correlation heatmap and a cumulative returns chart.                                                                             

Also, added few more prompt to fix the library issues and chart alignment, which will be seen at the end of the convo with chatgpt.

[ChatGPT](https://chatgpt.com/share/69e53ea2-e08c-83ea-bc5c-443398df8c46)

```{r}
############################################################
# ETF Return Analysis and Visualization
# Assets: SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, GLD
# Measures:
#  - Daily log returns
#  - Annualized returns
#  - Annualized volatility
#  - Correlation matrix
#  - Correlation heatmap
#  - Cumulative returns chart
############################################################

# ----------------------------------------------------------
# 1. Install required packages
# ----------------------------------------------------------
required_packages <- c("quantmod", "xts", "PerformanceAnalytics", "corrplot")

installed_packages <- rownames(installed.packages())

for (pkg in required_packages) {
  if (!(pkg %in% installed_packages)) {
    install.packages(pkg)
  }
}

# ----------------------------------------------------------
# 2. Load libraries
# ----------------------------------------------------------
library(quantmod)
library(xts)
library(PerformanceAnalytics)
library(corrplot)

# ----------------------------------------------------------
# 3. Define ETF tickers
# ----------------------------------------------------------
etf_symbols <- c("SPY", "QQQ", "IWM", "VEA", "VWO",
                 "AGG", "TLT", "LQD", "VNQ", "GLD")

# ----------------------------------------------------------
# 4. Set date range (last 3 years)
# ----------------------------------------------------------
end_date <- Sys.Date()
start_date <- end_date - 365 * 3

# ----------------------------------------------------------
# 5. Download ETF price data from Yahoo Finance
# ----------------------------------------------------------
getSymbols(Symbols = etf_symbols,
           src = "yahoo",
           from = start_date,
           to = end_date,
           auto.assign = TRUE)

# ----------------------------------------------------------
# 6. Extract Adjusted Closing Prices
# ----------------------------------------------------------
adjusted_prices_list <- lapply(etf_symbols, function(symbol) {
  Ad(get(symbol))
})

# Rename each column with ticker symbol
for (i in seq_along(adjusted_prices_list)) {
  colnames(adjusted_prices_list[[i]]) <- etf_symbols[i]
}

# Merge into one xts object
adjusted_prices <- do.call(merge, adjusted_prices_list)

# Preview price data
head(adjusted_prices)
tail(adjusted_prices)

# ----------------------------------------------------------
# 7. Calculate Daily Log Returns
# ----------------------------------------------------------
log_returns <- na.omit(Return.calculate(adjusted_prices, method = "log"))

# Preview return data
head(log_returns)
tail(log_returns)

# ----------------------------------------------------------
# 8. Calculate Annualized Return and Volatility
# ----------------------------------------------------------
annualized_returns <- apply(log_returns, 2, Return.annualized, scale = 252)
annualized_volatility <- apply(log_returns, 2, StdDev.annualized, scale = 252)

# Combine into one summary table
summary_stats <- data.frame(
  ETF = colnames(log_returns),
  Annualized_Return = as.numeric(annualized_returns),
  Annualized_Volatility = as.numeric(annualized_volatility)
)

print(summary_stats)

# Optional: format as percentages
summary_stats$Annualized_Return <- round(summary_stats$Annualized_Return * 100, 2)
summary_stats$Annualized_Volatility <- round(summary_stats$Annualized_Volatility * 100, 2)

print(summary_stats)

# ----------------------------------------------------------
# 9. Correlation Matrix
# ----------------------------------------------------------
correlation_matrix <- cor(log_returns, use = "complete.obs")

print(correlation_matrix)

# ----------------------------------------------------------
# 10. Correlation Heatmap
# ----------------------------------------------------------
corrplot(correlation_matrix,
         method = "color",
         type = "upper",
         order = "hclust",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         number.cex = 0.7,
         mar = c(0, 0, 1, 0),
         title = "Correlation Heatmap of Daily Log Returns")

# ----------------------------------------------------------
# 11. Cumulative Returns Chart
# ----------------------------------------------------------
chart.CumReturns(log_returns,
                 wealth.index = TRUE,
                 legend.loc = "topleft",
                 main = "Cumulative Returns of ETFs (Last 3 Years)")

# ----------------------------------------------------------
# 12. Optional: Save outputs
# ----------------------------------------------------------
write.csv(summary_stats,
          file = "etf_summary_statistics.csv",
          row.names = FALSE)

write.csv(correlation_matrix,
          file = "etf_correlation_matrix.csv",
          row.names = TRUE)
```

### 3RD AI Prompt

I want you to Write R code for a mean-variance optimized portfolio using PortfolioAnalytics and I want it to align to the results of my initial prompt  "https://chatgpt.com/share/69e53ea2-e08c-83ea-bc5c-443398df8c46"

[ChatGPT](https://chatgpt.com/share/69e54444-163c-83ea-9d5c-27a794c7aeac)


```{r}
############################################################
# Mean-Variance Optimized Portfolio using PortfolioAnalytics
# ETFs: SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, GLD
# Data Source: Yahoo Finance
# Horizon: Last 3 Years
############################################################

# =========================
# 1. Install and load packages
# =========================
required_packages <- c(
  "quantmod",
  "PerformanceAnalytics",
  "PortfolioAnalytics",
  "ROI",
  "ROI.plugin.quadprog",
  "ROI.plugin.glpk",
  "xts",
  "zoo",
  "ggplot2",
  "reshape2",
  "scales"
)

installed_packages <- rownames(installed.packages())

for (pkg in required_packages) {
  if (!(pkg %in% installed_packages)) {
    install.packages(pkg, dependencies = TRUE)
  }
}

library(quantmod)
library(PerformanceAnalytics)
library(PortfolioAnalytics)
library(ROI)
library(ROI.plugin.quadprog)
library(ROI.plugin.glpk)
library(xts)
library(zoo)
library(ggplot2)
library(reshape2)
library(scales)

# =========================
# 2. Define ETF universe and download data
# =========================
etfs <- c("SPY", "QQQ", "IWM", "VEA", "VWO", "AGG", "TLT", "LQD", "VNQ", "GLD")

start_date <- Sys.Date() - 365 * 3
end_date   <- Sys.Date()

getSymbols(Symbols = etfs,
           src = "yahoo",
           from = start_date,
           to = end_date,
           auto.assign = TRUE,
           warnings = FALSE)

# =========================
# 3. Extract Adjusted Close prices and merge
# =========================
price_list <- lapply(etfs, function(ticker) Ad(get(ticker)))
prices <- do.call(merge, price_list)
colnames(prices) <- etfs

# Remove rows with all missing values, then carry forward/backward if needed
prices <- prices[!apply(is.na(prices), 1, all), ]
prices <- na.locf(prices, na.rm = FALSE)
prices <- na.locf(prices, fromLast = TRUE)

# =========================
# 4. Compute daily log returns
# =========================
returns <- na.omit(Return.calculate(prices, method = "log"))
colnames(returns) <- etfs

# =========================
# 5. Summary statistics
# =========================
annualized_returns <- Return.annualized(returns, geometric = TRUE)
annualized_vol     <- StdDev.annualized(returns)
cor_matrix         <- cor(returns, use = "complete.obs")

cat("Annualized Returns:\n")
print(round(annualized_returns, 4))

cat("\nAnnualized Volatility:\n")
print(round(annualized_vol, 4))

cat("\nCorrelation Matrix:\n")
print(round(cor_matrix, 3))

# =========================
# 6. Build Mean-Variance Portfolio Specification
# =========================
portfolio_spec <- portfolio.spec(assets = colnames(returns))

# Full investment constraint
portfolio_spec <- add.constraint(
  portfolio = portfolio_spec,
  type = "full_investment"
)

# Long-only constraint
portfolio_spec <- add.constraint(
  portfolio = portfolio_spec,
  type = "long_only"
)

# Optional box constraint to avoid overly concentrated weights
portfolio_spec <- add.constraint(
  portfolio = portfolio_spec,
  type = "box",
  min = 0,
  max = 0.35
)

# Mean objective: maximize expected return
portfolio_spec <- add.objective(
  portfolio = portfolio_spec,
  type = "return",
  name = "mean"
)

# Risk objective: minimize portfolio variance
portfolio_spec <- add.objective(
  portfolio = portfolio_spec,
  type = "risk",
  name = "var"
)

# =========================
# 7. Optimize the portfolio
# =========================
opt_mv <- optimize.portfolio(
  R = returns,
  portfolio = portfolio_spec,
  optimize_method = "ROI",
  trace = TRUE
)

# Extract optimal weights
mv_weights <- extractWeights(opt_mv)

cat("\nMean-Variance Optimal Weights:\n")
print(round(mv_weights, 4))

# =========================
# 8. Portfolio returns
# =========================
mv_portfolio_returns <- Return.portfolio(
  R = returns,
  weights = mv_weights,
  rebalance_on = "months"
)

colnames(mv_portfolio_returns) <- "MeanVariance_Portfolio"

# =========================
# 9. Compare against Equal Weight portfolio
# =========================
equal_weights <- rep(1 / length(etfs), length(etfs))
names(equal_weights) <- etfs

ew_portfolio_returns <- Return.portfolio(
  R = returns,
  weights = equal_weights,
  rebalance_on = "months"
)

colnames(ew_portfolio_returns) <- "EqualWeight_Portfolio"

portfolio_compare <- na.omit(merge(mv_portfolio_returns, ew_portfolio_returns))

# =========================
# 10. Performance summary
# =========================
cat("\nPortfolio Performance Comparison:\n")

comparison_stats <- rbind(
  "Annualized Return" = Return.annualized(portfolio_compare, geometric = TRUE),
  "Annualized Volatility" = StdDev.annualized(portfolio_compare),
  "Sharpe Ratio" = SharpeRatio.annualized(portfolio_compare, Rf = 0, geometric = TRUE)
)

print(round(comparison_stats, 4))

# =========================
# 11. Efficient Frontier
# =========================
frontier <- create.EfficientFrontier(
  R = returns,
  portfolio = portfolio_spec,
  type = "mean-StdDev",
  n.portfolios = 25
)

chart.EfficientFrontier(frontier,
                        match.col = "StdDev",
                        n.portfolios = 25,
                        main = "Efficient Frontier")

# =========================
# 12. Cumulative returns chart
# =========================
charts.PerformanceSummary(
  portfolio_compare,
  main = "Mean-Variance vs Equal-Weight Portfolio"
)

# =========================
# 13. Correlation heatmap
# =========================
cor_df <- melt(cor_matrix)
colnames(cor_df) <- c("ETF1", "ETF2", "Correlation")

ggplot(cor_df, aes(x = ETF1, y = ETF2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "darkred",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Correlation Heatmap of ETF Daily Log Returns",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# =========================
# 14. Plot portfolio weights
# =========================
weights_df <- data.frame(
  ETF = names(mv_weights),
  Weight = as.numeric(mv_weights)
)

ggplot(weights_df, aes(x = reorder(ETF, Weight), y = Weight)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Mean-Variance Optimized Portfolio Weights",
    x = NULL,
    y = "Portfolio Weight"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# =========================
# 15. Optional: Display top-level summary neatly
# =========================
cat("\n=============================\n")
cat("MEAN-VARIANCE PORTFOLIO SUMMARY\n")
cat("=============================\n")

cat("\nOptimal Weights:\n")
print(round(mv_weights, 4))

cat("\nAnnualized Return (MV Portfolio): ",
    round(as.numeric(Return.annualized(mv_portfolio_returns, geometric = TRUE)), 4), "\n", sep = "")
cat("Annualized Volatility (MV Portfolio): ",
    round(as.numeric(StdDev.annualized(mv_portfolio_returns)), 4), "\n", sep = "")
cat("Sharpe Ratio (MV Portfolio): ",
    round(as.numeric(SharpeRatio.annualized(mv_portfolio_returns, Rf = 0, geometric = TRUE)), 4), "\n", sep = "")
```

### 4TH AI Prompt

Now, help me backtest the optimized portfolio against an equal-weight portfolio and a 60/40 benchmark

[ChatGPT](https://chatgpt.com/share/69e54735-17d8-83ea-a2b9-9c53fbf04aa9)

```{r}
############################################################
# STEP: Load Required Packages
############################################################

pkgs <- c(
  "quantmod",
  "PerformanceAnalytics",
  "PortfolioAnalytics",
  "ROI",
  "ROI.plugin.quadprog",
  "xts",
  "ggplot2",
  "reshape2"
)

new_pkgs <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs)

lapply(pkgs, library, character.only = TRUE)

############################################################
# STEP: Define ETFs and Download Adjusted Prices
############################################################

symbols <- c("SPY", "QQQ", "IWM", "VEA", "VWO", "AGG", "TLT", "LQD", "VNQ", "GLD")

start_date <- Sys.Date() - 365 * 3
end_date   <- Sys.Date()

getSymbols(symbols, src = "yahoo", from = start_date, to = end_date, auto.assign = TRUE)

############################################################
# STEP: Merge Adjusted Closing Prices
############################################################

prices <- do.call(merge, lapply(symbols, function(sym) Ad(get(sym))))
colnames(prices) <- symbols
prices <- na.omit(prices)

############################################################
# STEP: Calculate Daily Log Returns
############################################################

returns <- na.omit(Return.calculate(prices, method = "log"))

############################################################
# STEP: Mean-Variance Portfolio Optimization
############################################################

# Create portfolio specification
port_spec <- portfolio.spec(assets = colnames(returns))

# Add constraints
port_spec <- add.constraint(portfolio = port_spec, type = "full_investment")
port_spec <- add.constraint(portfolio = port_spec, type = "long_only")

# Add objectives
port_spec <- add.objective(portfolio = port_spec, type = "return", name = "mean")
port_spec <- add.objective(portfolio = port_spec, type = "risk", name = "var")

# Run optimization
opt_portfolio <- optimize.portfolio(
  R = returns,
  portfolio = port_spec,
  optimize_method = "ROI"
)

# Extract optimized weights
opt_weights <- extractWeights(opt_portfolio)

# Display optimized weights
opt_weights_df <- data.frame(
  ETF = names(opt_weights),
  Weight = round(as.numeric(opt_weights), 4)
)

print(opt_weights_df)

############################################################
# STEP: Construct Equal-Weight Portfolio
############################################################

n_assets <- ncol(returns)

weights_equal <- rep(1 / n_assets, n_assets)
names(weights_equal) <- colnames(returns)

ret_equal <- Return.portfolio(
  R = returns,
  weights = weights_equal,
  rebalance_on = "months"
)

############################################################
# STEP: Construct 60/40 Benchmark Portfolio
############################################################

# 60% allocated equally across equity ETFs
# 40% allocated equally across bond ETFs
# GLD is excluded from the traditional 60/40 benchmark

stock_assets <- c("SPY", "QQQ", "IWM", "VEA", "VWO", "VNQ")
bond_assets  <- c("AGG", "TLT", "LQD")

weights_6040 <- rep(0, n_assets)
names(weights_6040) <- colnames(returns)

weights_6040[stock_assets] <- 0.60 / length(stock_assets)
weights_6040[bond_assets]  <- 0.40 / length(bond_assets)

ret_6040 <- Return.portfolio(
  R = returns,
  weights = weights_6040,
  rebalance_on = "months"
)

############################################################
# STEP: Construct Mean-Variance Optimized Portfolio Returns
############################################################

ret_mvo <- Return.portfolio(
  R = returns,
  weights = opt_weights,
  rebalance_on = "months"
)

############################################################
# STEP: Merge Portfolio Returns
############################################################

portfolio_returns <- merge(ret_mvo, ret_equal, ret_6040)
colnames(portfolio_returns) <- c("Mean-Variance", "Equal Weight", "60/40")

portfolio_returns <- na.omit(portfolio_returns)

############################################################
# STEP: Performance Summary Tables
############################################################

cat("\nAnnualized Return Statistics:\n")
print(table.AnnualizedReturns(portfolio_returns))

cat("\nRisk Metrics:\n")
print(rbind(
  AnnualizedStdDev = StdDev.annualized(portfolio_returns),
  SharpeRatio = SharpeRatio.annualized(portfolio_returns, Rf = 0)
))

cat("\nMaximum Drawdown:\n")
print(maxDrawdown(portfolio_returns))

############################################################
# STEP: Cumulative Returns Chart
############################################################

chart.CumReturns(
  portfolio_returns,
  legend.loc = "topleft",
  main = "Cumulative Returns: Mean-Variance vs Equal Weight vs 60/40"
)

############################################################
# STEP: Performance Summary Chart
############################################################

charts.PerformanceSummary(
  portfolio_returns,
  main = "Performance Summary of Portfolio Strategies"
)

############################################################
# STEP: Correlation Heatmap of ETF Returns
############################################################

cor_mat <- cor(returns)
cor_df <- melt(cor_mat)

ggplot(cor_df, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(
    title = "Correlation Heatmap of ETF Daily Returns",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

############################################################
# STEP: Plot Optimized Portfolio Weights
############################################################

barplot(
  opt_weights,
  main = "Mean-Variance Optimized Portfolio Weights",
  col = "lightblue",
  las = 2
)
```

### 5TH AI Prompt

Create a cumulative returns chart and performance table using PerformanceAnalytics Make it the continuation of the discussion for the ETF setup: SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, and GLD over the last 3 years, using adjusted prices from Yahoo Finance. here's the link from our last convo 9backtesting portfolio strategies) https://chatgpt.com/share/69e54735-17d8-83ea-a2b9-9c53fbf04aa9

[ChatGPT](https://chatgpt.com/share/69e54aba-3c04-83ea-9416-b3bd9418670a)

```{r}
############################################################
# ETF Performance Analysis: Cumulative Returns + Performance Table
# ETFs: SPY, QQQ, IWM, VEA, VWO, AGG, TLT, LQD, VNQ, GLD
# Period: Last 3 Years
# Data Source: Yahoo Finance
############################################################

# ---------------------------
# 1. Install and load packages
# ---------------------------
pkgs <- c("quantmod", "PerformanceAnalytics", "xts")

new_pkgs <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs) > 0) install.packages(new_pkgs)

library(quantmod)
library(PerformanceAnalytics)
library(xts)

# ---------------------------
# 2. Define ETF tickers
# ---------------------------
tickers <- c("SPY", "QQQ", "IWM", "VEA", "VWO", "AGG", "TLT", "LQD", "VNQ", "GLD")

# ---------------------------
# 3. Download adjusted prices from Yahoo Finance
# ---------------------------
getSymbols(Symbols = tickers, src = "yahoo", from = Sys.Date() - 365*3, auto.assign = TRUE)

# ---------------------------
# 4. Merge adjusted closing prices into one xts object
# ---------------------------
prices_list <- lapply(tickers, function(ticker) Ad(get(ticker)))
prices_xts  <- do.call(merge, prices_list)
colnames(prices_xts) <- tickers

# Remove missing rows
prices_xts <- na.omit(prices_xts)

# ---------------------------
# 5. Calculate daily log returns
# ---------------------------
returns_xts <- na.omit(Return.calculate(prices_xts, method = "log"))

# ---------------------------
# 6. Cumulative returns chart
# ---------------------------
chart.CumReturns(
  returns_xts,
  wealth.index = TRUE,
  legend.loc = "topleft",
  main = "Cumulative Returns of ETFs Over the Last 3 Years",
  ylab = "Growth of $1",
  lwd = 2
)

# ---------------------------
# 7. Performance table using PerformanceAnalytics
# ---------------------------
# table.Stats gives a broad descriptive summary
performance_stats <- table.Stats(returns_xts)
print(round(performance_stats, 4))

# ---------------------------
# 8. Cleaner summary table with key portfolio metrics
# ---------------------------
annual_return <- Return.annualized(returns_xts)
annual_vol    <- StdDev.annualized(returns_xts)
sharpe_ratio  <- SharpeRatio.annualized(returns_xts, Rf = 0)
max_drawdown  <- maxDrawdown(returns_xts)

performance_summary <- data.frame(
  ETF = rownames(annual_return),
  Annualized_Return = round(as.numeric(annual_return[, 1]), 4),
  Annualized_Volatility = round(as.numeric(annual_vol[, 1]), 4),
  Sharpe_Ratio = round(as.numeric(sharpe_ratio[, 1]), 4),
  Max_Drawdown = round(as.numeric(max_drawdown[, 1]), 4)
)

rownames(performance_summary) <- NULL

cat("\n========================================\n")
cat("Key Performance Summary Table\n")
cat("========================================\n")
print(performance_summary)

# ---------------------------
# 9. Optional full performance dashboard
# ---------------------------
charts.PerformanceSummary(
  returns_xts,
  main = "Performance Summary of ETFs Over the Last 3 Years"
)
```

### 6TH AI Prompt

I want you to explain the limitations of an in-sample portfolio backtest in plain academic language.

[ChatGPT](https://chatgpt.com/share/69e54b62-7c80-83ea-a08e-859acd03f3a2)
